Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-03-06 Appendix_plain_v3.tex
Updates from Overleaf
Update manuscript and main text for submission.

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu L$ aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see \cite{shelton2022} Supp S1 for additional description). Here we fix $\zeta$ = \textbf{-1.15} based on previous analyses (\cite{shelton2022}).

@@

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu L^{-1}$, each of 2$\mu L$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{qp}$ be the known log-copy number for sample $q$ in PCR plate $p$, then,

@@

As above, we let $D_{sj}$ be the log-scale eDNA concentration but now for species $s$. We let $E_{sji}$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; \textbf{Aitchison 1986}).  We define a reference species in each sample, $R_i$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see \cite{shelton2023}).  We can then calculate $\nu_{sji}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see \textit{Mock Communities} below),

@@

\begin{align} 
	 log(F_s(x,y,d)) = \kappa + \eta(x,y) + \omega_{d}(x,y,d) \\
	 G_{s}(x,y,d) \sim \mathrm{Tweedie}(F_s(x,y,d),p ,\theta_s)
\end{align}
where x, y, and d index longitude, latitude, and depth, respectively.   $\kappa$ is scalar intercept. Both $\eta(x,y)$ and $\omega_d(x,y,d)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim  \mathrm{MVN}(0,\sum_\eta)$ and $\omega_d \sim  \mathrm{MVN}(0,\sum_{\omega_d})$ (see \cite{anderson2022}) where $\eta$ is a depth-invariant spatial field and $\omega_d$ is a spatial field specific to each depth category. In simple terms, each location-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated DNA, while $\omega_d$ modulates that estimate for different sampling depths. The $\omega_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$. Additionally to convert eDNA concentration from from copies/$\mu$L (reaction concentration) to copies/L of water filtered (ecological concentration) we multiply $F_s(x,y,d)$ by 20 which is the conversion value indicating the ratio of filtered volume and the volume of DNA template used in the PCR reaction.

@@

The results included in this article are based on the predicted smooth mean concentration, $F_s(x,y,d)$.

@@

&Parameters&\\
\hline
$\psi$& overall mean hake eDNA concentration in ln(copies/$\mu$L) & $\mathcal{N}(2,2)$\\
$\epsilon_{j}$& spatial and depth specific random effect & $\mathcal{N}(0,\tau)$\\
$\tau$& standard deviation of biological replicate random effect & $\mathcal{TN}(0,0.2;0, +\infty)$\\

$\delta$& biological replicate random effect & $\mathcal{N}(0,\sigma_D)$\\
$\sigma_D $& standard deviation of spatial and depth specific random effect & $\mathcal{TN}(0,3;0, +\infty)$\\

@@

    \end{tabular}
    \caption{Data, state processes, parameters, and subscripts employed in the joint Bayesian model and their prior distributions}
    \label{tab:priortable}
\end{table}

\clearpage
\begin{table}\centering
\caption{Species specific metabarcoding amplification efficiency relative to reference species (R; \textit{Merluccius productus} here).}
    \begin{tabular}{lc}
\textbf{Species} & $\alpha$ \\
\textit{Clupea pallasii} &   -0.00138 \\
\textit{Engraulis mordax} &    0.00557 \\
\textit{Leuroglossus stilbius} &   -0.00390 \\
\textit{Microstomus pacificus} &    0.00194 \\
\textit{Sardinops sagax} &    0.00705 \\
\textit{Scomber japonicus} &    0.00434 \\
\textit{Sebastes entomelas} &    0.00459 \\
\textit{Stenobrachius leucopsarus} &    0.00970 \\
\textit{Tactostoma macropus} &   -0.00091 \\
\textit{Tarletonbeania crenularis} &    0.00144 \\
\textit{Thaleichthys pacificus} &   -0.00506 \\
\textit{Trachurus symmetricus} &   -0.00875 \\
\textit{Merluccius productus} &    0.00000 \\
    \end{tabular}
        \label{tab:alphas}
\end{table}
\clearpage

>>>>

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu$L aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see \cite{shelton2022} Supp S1 for additional description). Here we fix $\zeta$ = -1.15 based on previous analyses (\cite{shelton2022}).

@@

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu \mathrm{L}^{-1}$, each of 2$\mu \mathrm{L}$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{qp}$ be the known log-copy number for sample $q$ in PCR plate $p$, then,

@@

As above, we let $D_{sj}$ be the log-scale eDNA concentration but now for species $s$. We let $E_{sji}$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; \cite{aitchison1986}).  We define a reference species in each sample, $R_i$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see \cite{shelton2023}).  We can then calculate $\nu_{sji}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see \textit{Mock Communities} below),

@@

\begin{align} 
	 log(F_s(x,y,d)) = \kappa + \eta(x,y) + \omega_{d}(x,y,d) \\
	 G_{s}(x,y,d) \sim \mathrm{Tweedie}(F_s(x,y,d),p ,\theta_s)
\end{align}
where $x$, $y$, and $d$ index longitude, latitude, and depth, respectively.   $\kappa$ is scalar intercept. Both $\eta(x,y)$ and $\omega_d(x,y,d)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim  \mathrm{MVN}(0,\sum_\eta)$ and $\omega_d \sim  \mathrm{MVN}(0,\sum_{\omega_d})$ (see \cite{anderson2022}) where $\eta$ is a depth-invariant spatial field and $\omega_d$ is a spatial field specific to each depth category. In simple terms, each location-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated DNA, while $\omega_d$ modulates that estimate for different sampling depths. The $\omega_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$. Additionally to convert eDNA concentration from from copies/$\mu$L (reaction concentration) to copies/L of water filtered (ecological concentration) we multiply $F_s(x,y,d)$ by 20 which is the conversion value indicating the ratio of filtered volume and the volume of DNA template used in the PCR reaction.

@@

All the results included in this article are based on the predicted smooth mean concentration, $F_s(x,y,d)$.

@@

&Parameters&\\
\hline
$\psi$& overall mean hake eDNA concentration in ln(copies/$\mu$L) & $\mathcal{N}(2,2)$\\
$\epsilon_{j}$& spatial and depth specific random effect & $\mathcal{N}(0,\tau)$\\
$\tau$& standard deviation of biological replicate random effect & $\mathcal{H-N}(0,0.2;0, +\infty)$\\

$\delta$& biological replicate random effect & $\mathcal{N}(0,\sigma_D)$\\
$\sigma_D $& standard deviation of spatial and depth specific random effect & $\mathcal{H-N}(0,3;0, +\infty)$\\

@@

    \end{tabular}
    \caption{Data, state processes, parameters, and subscripts employed in the joint Bayesian model and their prior distributions.  $N$ and $H-N$ designate a normal and half-normal distribution, respectively.}
    \label{tab:priortable}
\end{table}

\clearpage
\begin{table}\centering
\caption{Species specific metabarcoding amplification efficiency relative to reference species (R; \textit{Merluccius productus} (Pacific hake) here).}
    \begin{tabular}{lc}
\textbf{Species} & $\alpha$ \\
\textit{Clupea pallasii} &   -0.00138 \\
\textit{Engraulis mordax} &    0.00557 \\
\textit{Leuroglossus stilbius} &   -0.00390 \\
\textit{Microstomus pacificus} &    0.00194 \\
\textit{Sardinops sagax} &    0.00705 \\
\textit{Scomber japonicus} &    0.00434 \\
\textit{Sebastes entomelas} &    0.00459 \\
\textit{Stenobrachius leucopsarus} &    0.00970 \\
\textit{Tactostoma macropus} &   -0.00091 \\
\textit{Tarletonbeania crenularis} &    0.00144 \\
\textit{Thaleichthys pacificus} &   -0.00506 \\
\textit{Trachurus symmetricus} &   -0.00875 \\
\textit{Merluccius productus} &    0.00000 \\
    \end{tabular}
        \label{tab:alphas}
\end{table}
\clearpage

### 2025-03-06 Manuscript_plain_v3.tex
Update manuscript and main text for submission.

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and eAOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
All authors provided edits to the manuscript \\
Please check that this is correct and add/edit as needed XXX

@@

\section*{Abstract}
Molecular data from environmental samples can reflect the abundance of species’ DNA, an index immediately relevant to natural-resources management at a broad geographic scale. These data commonly derive from assays designed and targeted for specific species (e.g., using qPCR or dPCR), or metabarcoding assays from more general PCR primers that amplify many species simultaneously. Multispecies analyses make efficient use of field samples and laboratory time, and speak to inherently multispecies questions of management and ecological research. However, unlike single-species techniques, metabarcoding by itself can reflect only the proportions of target-species DNA present, not their absolute quantities. Here, we combine qPCR and metabarcoding data derived from the same samples to map quantities of many fish species along the US West Coast in three dimensions, demonstrating a technique of practical importance for both management and ecology. We derive spatially explicit estimates of eDNA abundance for 12 common species of ecological and management importance and point the way to quantitative surveys of wild species using molecules alone. We find that species distribution maps derived from eDNA effectively capture species niches and spatial distributions. Notably, our analyses identified biodiversity hotspots that align with previously documented regions of ecological significance, such as the Columbia River plume and the Heceta Bank.

@@

While trawling and other traditional sampling methods have been the fundamental tools for directly measuring fish abundance and biomass, they are often impractical for species inhabiting inaccessible habitats, occurring at low densities, or fragile species easily injured during capture. By contrast, eDNA sampling captures information on a broader range of taxa, including those that are difficult or impossible to study with traditional gear. As such, eDNA can meaningfully complement traditional stock assessment methods for commercially important species while at the same time revealing the abundance and distribution of elusive or understudied species.

@@

\section*{Results}
We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

Below we highlight results for six species (Fig. \ref{fig:fig1}) representing a range of habitat associations and general species information; distribution maps for all the remaining species are presented in \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}.

@@

\subsection*{Mid-water}
Widow rockfish (\textit{Sebastes entomelas}) is common between 100-350m between northern Baja California and southern Alaska, with pelagic larvae and juveniles generally detected in shallower habitats ($<$ 150m depth) \cite{reynolds2001,bosley2014} including in nearshore kelp forests \cite{quigley2024, laidig2007} . Moreover, because it is a common bycatch species in the Pacific hake (\textit{Merluccius productus}) fishery, concerns about bycatch of widow rockfish can also influence the much larger hake fishery (i.e., it is a “choke” species for hake; \cite{somers2018}). Despite its commercial importance, no targeted survey effort exists for widow rockfish, in part because bottom trawls do not sample its pelagic habitat well. We find widow rockfish eDNA strongly and nearly exclusively associated with 50m sample depths (Fig. \ref{fig:fig1} and \ref{fig:fig4}), consistent with the pelagic larvae and juveniles reported in the literature and somewhat shallower than adults are commonly found \cite{reynolds2001,bosley2014,quigley2024}, potentially reflecting the species’ nocturnal feeding schooling behavior \cite{reynolds2001,bosley2014}.

@@

\subsection*{Benthic}
Among deeper-water fishes, eDNA revealed spatial and potentially ontogenetic migration patterns for species that are common, yet under studied, such as the dragonfish (\textit{Tactostoma macropus}). Available literature suggests dragonfish occur roughly between 40$^\circ$N-50$^\circ$N (adults at 300-900m, larvae at 0-60m in summer) \cite{kawaguchi1993} and perform diel vertical migrations to around 100 meters \cite{kawaguchi1993,willis1982}. The eDNA concentrations appeared to reflect this activity (Fig. \ref{fig:fig4}), with the predominant surface signal at northern latitudes likely reflecting larvae and juveniles (Fig. \ref{fig:fig1}), and two deeper hotspots at 300m and 500m separated by about eight degrees of latitude. However, no other quantitative estimates of abundance or detailed distribution patterns for dragonfish exist in the current literature.

@@

\subsection*{eDNA as Management Tool}
By mapping three-dimensional spatial patterns, eDNA provides information on species distributions, habitat preferences, and ecological processes that is often otherwise unavailable. A single water sample can yield quantitative information on tens to hundreds of species -- including for species for which traditional fisheries surveys are not currently conducted -- without the need to develop species-specific assays. We find species-level eDNA concentrations align closely with the preferred habitats' physiological constraints (e.g., temperature tolerance or food availability) described for each species in the available literature. The observed latitudinal differences of eDNA concentration among species – such as those between anchovy and sardine – highlights species’ thermal ranges and largely coincide with other surveys \cite{zwolinski2024}; moreover, eDNA patterns appear to reflect diel vertical migrations as observed in myctophids, where eDNA concentrations are consistent with the species’ ontogenetic movement between surface and mid- to deep-water habitats. We also identify aggregations of richness and total eDNA near high-productivity regions of ecological and commercial value. Regions like Heceta Bank host dense aggregations of forage fish and are critical for both fishery management and for broader marine food-web stability.

@@

Moreover, integrating retrospective eDNA analyses into climate change modeling can enrich predictive frameworks by providing historically grounded baselines of community composition, thereby enhancing our understanding of ecosystem responses to environmental shifts. For example, \cite{diaz2020} demonstrated how archived samples could be leveraged to reconstruct past fish communities using eDNA, thereby complementing traditional monitoring methods and further extending the utility of archived material. Such retrospective applications deepen and extend the value of every research expedition, ensuring that once-collected samples continue to yield valuable ecological insights over time. We therefore see eDNA as an important datastream that maximizes the information value of existing research cruises, rather than a means of replacing those cruises themselves.

We have shown an example of the substantial value of multi-species eDNA analysis for fisheries and other natural-resource management questions \cite{ledger2024,stoeckle2024}, however, molecular patterns alone are unlikely to drive quantitative natural-resources decisions, in part because eDNA generally does not provide auxiliary information important for managing populations (e.g., information on size- or age structure, fecundity, sex, or disease). Moreover, the eDNA sampling we describe here depends upon survey vessels that can reach and sample from deep water – survey vessels that exist to carry out traditional acoustic or trawl-based surveys. Furthermore, eDNA survey provide data in units of eDNA concentration, not more familiar units of mass or abundance. Relating observed eDNA concentrations to species mass requires data external to the eDNA signal \cite{guri2024a}, and in particular, requires that we understand proportionalities between eDNA concentration and biomass that may vary among species. We therefore see eDNA analysis as strengthening and extending existing surveys, with different data streams complementing one another.

>>>>

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and AOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
All authors provided edits to the manuscript \\

@@

\section*{Abstract}
Molecular data from environmental samples can reflect the abundance of species’ DNA, an index immediately relevant to natural-resources management at a broad geographic scale. These data commonly derive from assays designed and targeted for specific species (e.g., using qPCR or dPCR), or metabarcoding assays from more general PCR primers that amplify many species simultaneously. Multispecies analyses make efficient use of field samples and laboratory time, and speak to inherently multispecies questions of management and ecological interest. However, unlike single-species techniques, metabarcoding by itself can reflect only the proportions of target-species DNA present, not their absolute quantities. Here, we combine qPCR and metabarcoding data derived from the same samples to map quantities of many fish species along the US West Coast in three dimensions, demonstrating a technique of practical importance for both management and ecology. We derive spatially explicit estimates of eDNA abundance for 12 common species of ecological and management importance and point the way to quantitative surveys of wild species using molecules alone. We find that species distribution maps derived from eDNA effectively capture species niches and spatial distributions. Notably, our analyses identified biodiversity hotspots that align with previously documented regions of ecological significance, such as the Columbia River plume and the Heceta Bank.

@@

Surveys of marine environments play a fundamental role in supporting fisheries and conservation of protected species but are often difficult and expensive endeavors. While trawling, acoustics, or other traditional sampling methods have long been used to directly measure species abundance and biomass, they can be impractical for species inhabiting inaccessible habitats, occurring at low densities, or for fragile species easily injured during capture. By contrast, eDNA sampling captures information on a broader range of taxa, including those that are difficult or impossible to study with traditional gear. As such, eDNA can meaningfully complement traditional stock assessment methods for commercially important species while at the same time revealing the abundance and distribution of elusive or understudied species.

@@

\section*{Results}

Previous eDNA work has tended to feature more qualitative multispecies results, or else quantitative analysis of single species at small spatial scales. Here, we create a rigorous quantitative framework and use hundreds of water samples to make ecological inferences at the scale of management. We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species -- which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} -- molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

%We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

Below we highlight results for six species (Fig. \ref{fig:fig1}) representing a range of habitat associations and general species information; distribution maps for all the remaining species are presented in \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}. In particular, our analyses present detailed, depth-specific information on abundance that is difficult to obtain with traditional sampling methods.

@@

\subsection*{Midwater}
Widow rockfish (\textit{Sebastes entomelas}) is common between 100-350m between northern Baja California and southern Alaska, with pelagic larvae and juveniles generally detected in shallower habitats ($<$ 150m depth) \cite{reynolds2001,bosley2014} including in nearshore kelp forests \cite{quigley2024, laidig2007} . Moreover, because it is a common bycatch species in the Pacific hake (\textit{Merluccius productus}) fishery, concerns about bycatch of widow rockfish can also influence the much larger hake fishery (i.e., it is a “choke” species for hake; \cite{somers2018}). Despite its commercial importance, no targeted survey effort exists for widow rockfish, in part because bottom trawls do not sample its pelagic habitat well. We find widow rockfish eDNA strongly and nearly exclusively associated with 50m sample depths (Fig. \ref{fig:fig1} and \ref{fig:fig4}), consistent with the pelagic larvae and juveniles reported in the literature and somewhat shallower than adults are commonly found \cite{reynolds2001,bosley2014,quigley2024}, potentially reflecting the species’ nocturnal feeding schooling behavior \cite{reynolds2001,bosley2014}.

@@

\subsection*{Deepwater}
Among deeper-water fishes, eDNA revealed spatial and potentially ontogenetic migration patterns for species that are common, yet under studied, such as the dragonfish (\textit{Tactostoma macropus}). Available literature suggests dragonfish occur roughly between 40$^\circ$N-50$^\circ$N (adults at 300-900m, larvae at 0-60m in summer) \cite{kawaguchi1993} and perform diel vertical migrations to around 100 meters \cite{kawaguchi1993,willis1982}. The eDNA concentrations appeared to reflect this activity (Fig. \ref{fig:fig4}), with the predominant surface signal at northern latitudes likely reflecting larvae and juveniles (Fig. \ref{fig:fig1}), and two deeper hotspots at 300m and 500m separated by about eight degrees of latitude. However, no other quantitative estimates of abundance or detailed distribution patterns for dragonfish exist in the current literature.

@@

\subsection*{eDNA as Management Tool}
By mapping three-dimensional spatial patterns, eDNA provides information on species distributions, habitat preferences, and ecological processes that is often otherwise unavailable. A single water sample can yield quantitative information on tens to hundreds of species -- including for species for which traditional fisheries surveys are not currently conducted -- without the need to develop species-specific assays. We find species-level eDNA concentrations align closely with the preferred habitats' physiological constraints (e.g., temperature tolerance or food availability) described for each species in the available literature. The observed latitudinal differences of eDNA concentration among species – such as those between anchovy and sardine – highlights species’ thermal ranges and largely coincide with other surveys \cite{zwolinski2024}; moreover, eDNA patterns appear to reflect diel vertical migrations as observed in myctophids, where eDNA concentrations are consistent with the species’ ontogenetic movement between surface and mid- to deep-water habitats. We also identify aggregations of richness and total eDNA near high-productivity regions of ecological and commercial value.

%Regions like Heceta Bank host dense aggregations of forage fish and are critical for both fishery management and for broader marine food-web stability.

@@

Moreover, integrating retrospective eDNA analyses a changing climate can enrich predictive frameworks by providing historically grounded baselines of community composition, thereby enhancing our understanding of ecosystem responses to environmental shifts. For example, \cite{diaz2020} demonstrated how archived samples could be leveraged to reconstruct past fish communities using eDNA, thereby complementing traditional monitoring methods and further extending the utility of archived material. Such retrospective applications deepen and extend the value of every research expedition, ensuring that once-collected samples continue to yield valuable ecological insights over time. We therefore see eDNA as an important datastream that maximizes the information value of existing research cruises, rather than a means of replacing those cruises themselves.

We have shown an example of the potential value of multi-species eDNA analysis for fisheries and other natural-resource management questions \cite{ledger2024,stoeckle2024}, however, molecular patterns alone are unlikely to drive quantitative natural-resources decisions, in part because eDNA generally does not provide auxiliary information important for managing populations (e.g., information on size- or age structure, fecundity, sex, or disease). Relating observed eDNA concentrations to species mass requires data external to the eDNA signal \cite{guri2024a}, and in particular, requires that we understand proportionalities between eDNA concentration and biomass that may vary among species. We therefore see eDNA analysis as strengthening and extending existing surveys, with different data streams complementing one another.

%Moreover, the eDNA sampling we describe here depends upon survey vessels that can reach and sample from deep water – survey vessels that exist to carry out traditional acoustic or trawl-based surveys. Furthermore, eDNA surveys provide data in units of eDNA concentration, not more familiar units of mass or abundance.
